xolotl: An Intuitive and Approachable Neuron \& Network Simulator

### 2018-08-14 paper/paper.tex
getting close to the final stretch

\title[xolotl: neuron \& network simulator]{Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
\topic{An Intuitive Neuronal Simulator}

@@

	Conductance-based models of neurons have been used extensively in computational neuroscience. However, working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in "C++" to objects in "MATLAB". Our approach builds on the tradition of uniting the speed and power of languages like "C++" with the ease-of-use and feature-set of scientific programming languages like "MATLAB". Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

	\tiny
	\keyFont{ \section{Keywords:} simulator, code:MATLAB, code:C++, conductance-based, neuron, network, pedagogy}

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of electrophysiology, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, each with several parameters. Second, many or all equations in these models can be strongly coupled through dynamical variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment can be different, and are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like "C++" or "FORTRAN" can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like "Python", "Julia", or "MATLAB".

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are written in a fast compiled language like "C" and allow for the construction and simulation of neuron models using object-oriented paradigms. These simulators tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python"  or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), and "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as "C" or "C++" (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, that we call "cpplab", which binds "MATLAB" code to classes specified in "C++" header files. This architecture automatically creates objects in "MATLAB" that represent the underlying object-oriented "C++" code, allowing the symbolic manipulation of "C++" objects in the "MATLAB" interface. In this paper, we introduce "xolotl", an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of "MATLAB" code using a hierarchical and intuitive syntax. Since models in the "MATLAB" workspace are automatically linked to models in the "C++" implementation, configuring these objects in "MATLAB" transparently configures the underlying "C++" objects. "xolotl" comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

"xolotl" is designed to be easy-to-use and richly featured while being fast enough to use in everyday research. Our focus was on designing an approachable simulator of conductance-based neurons and networks of these neurons; simulating arbitrary dynamical systems is therefore beyond the scope of this software. Specifically, the software was designed to simulate models of the form

@@

This software is designed to be used from within "MATLAB", a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make "xolotl" completely usable entirely from within "MATLAB". Models created using this simulator appear in the "MATLAB" workspace as native objects, are are thus fully scriptable, and are fully compatible with the large library of toolboxes that "MATLAB" provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new "C++" files on-the-fly. Finally, "xolotl" fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{A rich library of network components.} "xolotl" comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, Calcium level, and set of constituent components, representing either entire cells or parts of cells. Objects of type "conductance" represent populations of ion channels in a compartment that produce transmembrane currents. "synapse" objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type "mechanism" can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

\paragraph{Object-oriented.}
"xolotl" is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a "xolotl" model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the "xolotl" model tree (compartments, conductances, etc.) are bonafide "MATLAB" objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{Automatic type system.}
To circumvent the tradeoff between high-performance but hard to use languages like "C++" and richly-featured but potentially slow languages like "MATLAB", we constructed an automatic type system that links object oriented code in "C++" to object oriented code in "MATLAB". This architecture lets us construct the core of the simulator in "C++", leveraging features of "C++" like pointers that are not readily available in "MATLAB". A rudimentary way to make this "C++" code useable in "MATLAB" would be to re-write that code in "MATLAB" so that "MATLAB" objects can be bound to their "C++" implementations. However, this approach is cumbersome and inefficient, and can introduce errors. Instead, our automatic type system creates objects in "MATLAB" on-the-fly from "C++" class specifications, obviating the need to rewrite code in "MATLAB" while preserving a tight coupling between objects in the "MATLAB" workspace and their "C++" implementation. Crucially, this method makes developing new code much easier and simplifies the task of constructing complex frameworks that span these two languages.

\paragraph{Automatic hashing and compiling.}
Since every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. "xolotl" solves this problem by hashing (\cite{rivestMD5MessageDigestAlgorithm1992}) the "C++" header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its ``hash''). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the "MATLAB" workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

Our focus on ease-of-use and speed means some features were intentionally neglected in the design process.

\paragraph{Limited to conductance-based models.} "xolotl" has been developed specifically for conductance-based models. It does not currently support rate- or current-based models, or arbitrary dynamical systems.

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been shown to perform well in neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new network components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new "C++"  code. Currently, "xolotl" can only implement integration schemes with fixed step size.

\paragraph{Inefficient tools for handling large networks.} "xolotl" was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, "xolotl" is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new "C++" code. A new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, dayanTheoreticalNeuroscience2001}) requires creating a new "C++" header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying "C++" core code.

@@

In this section, we illustrate how "xolotl" can be used to generate, inspect, and simulate a variety of models. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as templates upon which researchers and educators can build.

@@

"xolotl" can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). Since the compartment is being voltage clamped, integrating the model returns the clamping current:

@@

By repeating the integration at a number of clamping voltages, we observe that the asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). "xolotl" can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. "xolotl" makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

A key intracellular mechanisms is the cytosolic buffering of Calcium and its influx through voltage-gated Calcium channels. Figure \ref{fig:figuremechanism}A shows a model of a single-compartment model with 8 populations of ion channels (\cite{liuModelNeuronActivityDependent1998}). Without any explicit mechanism for Calcium influx or buffering, the intracellular Calcium levels in this model do not change (Figure \ref{fig:figuremechanism}B) and the model tonically spikes (Figure \ref{fig:figuremechanism}C). Calcium buffering and influx can be modelled by a differential equation that increases intracellular Calcium with the current through Calcium channels and relaxes back to a baseline value (\cite{liuModelNeuronActivityDependent1998, prinzAlternativeHandtuningConductancebased2003, dayanTheoreticalNeuroscience2001}) (Figure \ref{fig:figuremechanism}D). This mechanism exists in the "xolotl" library as "CalciumMech1" and can be added to the model using a simple "add" statement (Figure \ref{fig:figuremechanism}E). The intracellular Calcium in the model now oscillates periodically (Figure \ref{fig:figuremechanism}F), synchronized to bursts in action potentials in this cell (Figure \ref{fig:figuremechanism}G).

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how "xolotl" can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, our simulator can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeller's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations. (\cite{deschutterConsumerGuideNeuronal1992}). "xolotl" makes it easy to bookmark configurations of a model and return to them at will. Internally, "xolotl" uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying "C++" implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into "xolotl" leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

@@

Network models in "xolotl" consist of "compartment" objects that can be connected by "synapse" objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use

@@

Our simulator has several types of synapses like Cholinergic, Glutamatergic, AMPA and NMDA built-in, and other synapse classes can be easily added using templates. Figure \ref{fig:figurenetwork} demonstrates the implementation of a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans (\cite{prinzSimilarNetworkActivity2004}). The pyloric model contains three compartments ("AB", "LP", and "PY") and seven synapses (Figure \ref{fig:figurenetwork}A). This structure is recapitulated in the hierarchy of the "xolotl" object, where conductances are contained within compartments (Figure \ref{fig:figurenetwork}B). The membrane potentials show triphasic rhythmicity in the three compartments (Figure \ref{fig:figurenetwork}C-E). When the "PY" is depolarized, the dynamical variable mediating the  glutamatergic ("Glut") synapse model between "PY" and "LP" is close to 1 and "LP" is inhibited (Figure \ref{fig:figurenetwork}F). Conversely, when "PY" is hyperpolarized, the dynamical variable is close to 0. When "PY" spikes, IPSPs can be seen in the "LP" voltage trace (Figure \ref{fig:figurenetwork}D-E).

@@

The "manipulate" function opens the GUI which permits visualization of changing parameters in real-time. Moving sliders representing the values of network parameters updates a plot (Figure \ref{fig:figuremanipulate}B). By default, the function opens a figure displaying the results of the "plot" function, which shows the voltage and intracellular calcium traces for each compartment (Figure \ref{fig:figuremanipulate}A). "manipulate" grants slider control over all "xolotl" parameters by default, but specific ones can be selected by passing them as arguments.

>>>>

\title[xolotl: neuron \& network simulator]{xolotl: An Intuitive and Approachable Neuron \& Network Simulator}
\author[\firstAuthorLast ]{\Authors}
\address{}
\correspondance{}
\extraAuth{}
\topic{An Intuitive Neuronal Simulator}

@@

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed and power of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Our software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. Our simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Our simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code promote reproducible and auditable research. Finally, our simulator is written in a modular fashion and has been released under a permissive free software license, enabling seamless integration into the workflow of other researchers.

	\tiny
	\keyFont{ \section{Keywords:} simulator, code:MATLAB, code:C++, conductance-based, neuron, network, neuron models}

@@

Nervous systems process and transmit information using electrically excitable membranes. Conductance-based models are a powerful biophysical simplification of an electrically excitable compartment in a neuron (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small circuits (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). These models provide an approachable framework for understanding many salient principles of electrophysiology, since they explicitly model cell membranes and ion channels as electrical components in a circuit. However, several challenges remain in working with biophysically-detailed conductance-based neuron models. First, these models can be high-dimensional with many nonlinear differential equations, each with several parameters. Second, many or all equations in these models can be strongly coupled through dynamical variables like the membrane potential.  In multi-compartment models of spatially extended neurons, membrane potentials in every compartment can be different, and are coupled to each other. Finally, the choice of programming language used to implement the model imposes tradeoffs in designing and using neuron and network simulators: simulators written in languages like C++ or FORTRAN can integrate equations quickly, but often lack the ease-of-use and interoperability of those written in scientific programming languages like Python, Julia, or MATLAB (Mathworks).

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like "Python"  or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}), "morphforge" (\cite{hullMorphforgeToolboxSimulating2014}), and "PyNN" (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as "C" or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Since models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

To overcome these design limitations, we have developed a novel automatic type system, that we call "cpplab", which binds MATLAB code to classes specified in C++ header files. This architecture automatically creates objects in MATLAB that represent the underlying object-oriented C++ code, allowing the symbolic manipulation of C++ objects in the MATLAB interface. In this paper, we introduce xolotl, an implementation of the "cpplab" system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of MATLAB code using a hierarchical and intuitive syntax. Since models in the MATLAB workspace are automatically linked to models in the C++ implementation, configuring these objects in MATLAB transparently configures the underlying C++ objects. xolotl comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

xolotl is designed to be easy-to-use and richly featured while being fast enough to use in everyday research. Our focus was on designing an approachable simulator of conductance-based neurons and networks of these neurons; simulating arbitrary dynamical systems is therefore beyond the scope of this software. Specifically, the software was designed to simulate models of the form

@@

This software is designed to be used from within MATLAB, a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make xolotl completely usable entirely from within MATLAB. Models created using this simulator appear in the MATLAB workspace as native objects, are are thus fully scriptable, and are fully compatible with the large library of toolboxes that MATLAB provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new C++ files on-the-fly. Finally, xolotl fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{A rich library of network components.} xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, Calcium level, and set of constituent components, representing either entire cells or parts of cells. Objects of type "conductance" represent populations of ion channels in a compartment that produce transmembrane currents. "synapse" objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type "mechanism" can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

\paragraph{Object-oriented.}
xolotl is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a xolotl model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the xolotl model tree (compartments, conductances, etc.) are bonafide MATLAB objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{Automatic type system.}
To circumvent the tradeoff between high-performance but hard to use languages like C++ and richly-featured but potentially slow languages like MATLAB, we constructed an automatic type system that links object oriented code in C++ to object oriented code in MATLAB. This architecture lets us construct the core of the simulator in C++, leveraging features of C++ like pointers that are not readily available in MATLAB. A rudimentary way to make this C++ code useable in MATLAB would be to re-write that code in MATLAB so that MATLAB objects can be bound to their C++ implementations. However, this approach is cumbersome and inefficient, and can introduce errors. Instead, our automatic type system creates objects in MATLAB on-the-fly from C++ class specifications, obviating the need to rewrite code in MATLAB while preserving a tight coupling between objects in the MATLAB workspace and their C++ implementation. Crucially, this method makes developing new code much easier and simplifies the task of constructing complex frameworks that span these two languages.

\paragraph{Automatic hashing and compiling.}
Since every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. xolotl solves this problem by hashing (\cite{rivestMD5MessageDigestAlgorithm1992}) the C++ header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its ``hash''). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the MATLAB workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

Our focus on our simulator's ease-of-use and speed imposed some limitations on its feature set.

\paragraph{Limited to conductance-based models.} xolotl has been developed specifically for conductance-based models. It does not currently support rate- or current-based models, or arbitrary dynamical systems.

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been shown to perform well in neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new network components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new C++  code. Currently, xolotl can only implement integration schemes with fixed step size.

\paragraph{Inefficient tools for handling large networks.} xolotl was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, xolotl is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

\paragraph{New mechanisms require new \texttt{C++} code.} Adding new network components requires writing new C++ code. A new conductance in the Hodgkin-Huxley formalism (\cite{hodgkinComponentsMembraneConductance1952, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, dayanTheoreticalNeuroscience2001}) requires creating a new C++ header file, though this is generally trivial. Implementing a new integration scheme or component type requires much more in-depth knowledge of the underlying C++ core code.

@@

In this section, we illustrate how xolotl can be used to generate, inspect, and simulate a variety of models. These examples have been chosen to illustrate various features of xolotl, and are intended to serve as templates upon which researchers and educators can build.

@@

xolotl can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). Since the compartment is being voltage clamped, integrating the model returns the clamping current:

@@

By repeating the integration at a number of clamping voltages, we observe that the asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). xolotl can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. xolotl makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

A key intracellular mechanisms is the cytosolic buffering of Calcium and its influx through voltage-gated Calcium channels. Figure \ref{fig:figuremechanism}A shows a model of a single-compartment model with 8 populations of ion channels (\cite{liuModelNeuronActivityDependent1998}). Without any explicit mechanism for Calcium influx or buffering, the intracellular Calcium levels in this model do not change (Figure \ref{fig:figuremechanism}B) and the model tonically spikes (Figure \ref{fig:figuremechanism}C). Calcium buffering and influx can be modelled by a differential equation that increases intracellular Calcium with the current through Calcium channels and relaxes back to a baseline value (\cite{liuModelNeuronActivityDependent1998, prinzAlternativeHandtuningConductancebased2003, dayanTheoreticalNeuroscience2001}) (Figure \ref{fig:figuremechanism}D). This mechanism exists in the xolotl library as "CalciumMech1" and can be added to the model using a simple "add" statement (Figure \ref{fig:figuremechanism}E). The intracellular Calcium in the model now oscillates periodically (Figure \ref{fig:figuremechanism}F), synchronized to bursts in action potentials in this cell (Figure \ref{fig:figuremechanism}G).

Neurons can regulate their electrical activity by controlling the spectrum of ion channels they express (\cite{macleanActivityIndependentHomeostasisRhythmically2003, turrigianoSelectiveRegulationCurrent1995, schulzVariableChannelExpression2006}). Here, we will show how xolotl can be used to represent a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration (\cite{olearyCorrelationsIonChannel2013, olearyCellTypesNetwork2014}) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, an object corresponding to this mechanism is added to each "conductance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target Calcium concentration (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from quiescence to truncated bursts of action potentials to periodic bursting as this mechanism regulates the neuron's ion channel spectrum. In summary, our simulator can be used to construct neuron models with intracellular mechanisms such as Calcium influx and buffering, and homeostatic regulation.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeller's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations. (\cite{deschutterConsumerGuideNeuronal1992}). xolotl makes it easy to bookmark configurations of a model and return to them at will. Internally, xolotl uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying C++ implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into xolotl leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

@@

Network models in xolotl consist of "compartment" objects that can be connected by "synapse" objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, to connect two single-compartment neurons called "LP" and "PY" using a chemical synapse of type "Cholinergic" with strength 30 nS, we can use

@@

Our simulator has several types of synapses like Cholinergic, Glutamatergic, AMPA and NMDA built-in, and other synapse classes can be easily added using templates. Figure \ref{fig:figurenetwork} demonstrates the implementation of a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans (\cite{prinzSimilarNetworkActivity2004}). The pyloric model contains three compartments ("AB", "LP", and "PY") and seven synapses (Figure \ref{fig:figurenetwork}A). This structure is recapitulated in the hierarchy of the xolotl object, where conductances are contained within compartments (Figure \ref{fig:figurenetwork}B). The membrane potentials show triphasic rhythmicity in the three compartments (Figure \ref{fig:figurenetwork}C-E). When the "PY" is depolarized, the dynamical variable mediating the  glutamatergic ("Glut") synapse model between "PY" and "LP" is close to 1 and "LP" is inhibited (Figure \ref{fig:figurenetwork}F). Conversely, when "PY" is hyperpolarized, the dynamical variable is close to 0. When "PY" spikes, IPSPs can be seen in the "LP" voltage trace (Figure \ref{fig:figurenetwork}D-E).

@@

Conductance-based neuron models are typically high dimensional and contain several parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.
